Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-01-22 draft.tex
[draft] comments on sec. III and appendix

In several interesting processes we might only have an approximate
result for $P$.  In this section, we focus on one particular class of
mismodelling of the signal PDF: the angular momentum-cutoff in partial
wave expansions. This mismodelling potentially affects a large number
of decays and scattering processes. For the sake of clarity we take
the interesting class of four-body decays $B\to P_1 P_2 \ell_1
\bar\ell_2$ \fred{How general is this? The plot implies that the
  dimeson and the dilepton do not have a common vertex. But this
  requires an intermediate state, e.g., $K^{*}$}as an
example\footnote{This includes the rare $b\to s$ mediated $B$ decay $B
  \to K\pi\ell^+\ell^-$, and the $V_{ub}$ suppressed decay $B\to
  \pi\pi\ell^+\bar\nu_\ell$. For both examples the PDF $P$ is known in
  the small-width approximation and when assuming a pure $P$-wave,
  resonant final state. An extension to $S-P$ interference has been
  studied for $B\to K\pi\ell^+\ell^-$
  \cite{Blake:2012mb,Becirevic:2011bp}, and for $\bar{B}\to
  \pi\pi\ell^-\bar\nu_\ell$ \cite{Faller:2013dwa}. For a first study
  of $S$, $P$
  and $D$ interference, see \cite{Das:2014sra}}.\\

@@

\subsection{Recipe for the Determination of Detector Effects}
\label{sec:systematics:acceptance}

@@

In order to finally extract the angular observables from data, we use the \emph{measured raw moments}. Their MC estimator
$\lbrace \est{Q}_i\rbrace$ -- based on the \emph{detected events} $x^{(n)}$, $n=1,\dots,N$ -- reads
\begin{equation}
    \est{Q}_i \equiv R \frac{1}{N} \sum_n^N \tilde{f}_i(x^{(n)})\,.
\end{equation}
We then obtain MC estimators of the angular observables via
\begin{equation}
    \label{eq:estS}
    \est{S}_i \equiv \sum_{i} \left(\hat{M}^{-1}\right)_{ij} \est{Q}_j\,.
\end{equation}
The precise value of $R$ (see \refeq{def-R}) depends on the a priori
unknown angular observables $\lbrace S_i\rbrace$.  However, ratios and
other suitable combinations of the estimators $\lbrace
\est{S}_i\rbrace$ are independent of $R$. Conversely, $R$ can be
extracted from the self-consistency condition\footnote{This condition
  is process dependent; see \refeq{recipe:lambdabtolambdall} and
  \refeq{recipe:btokstarll}.}  $\hat{S}_0 = \frac{1}{2}$. Since $R$ is
relevant in the determination of the branching ratio, it is therefore
useful to carry out that determination and the extraction of the
angular observables in a combined analysis.

@@

\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/error-Q2_5_6_S5.png}
            \includegraphics[width=0.45\textwidth]{figs/error-Q2_5_6_S7.png}
            \caption{Uncertainty of the angular observables $S_5$ and $S_7$, extracted from simulated events of the decay $B\to K^*\ell^+\ell^-$ and as a function of the number of simulated events $K$. The red curve represents a fit to the function given in \refeq{unc-on-mean}.}
        \label{fig:errors}
\end{figure}

@@

Let us assume that the angular decomposition has been achieved for some
PDF $P$ for all angles except one, which we label $\theta$. We now focus on only of of the resulting
observables, which we label $S(\theta) \equiv S(\vec{\nu},\theta)$. The dependence on the non-angular
variables will be suppressed in the following. \danny{Fred: is this less confusing now?}
Further, let $S$ have an expansion in terms of partial waves
$l_1, l_2 = 0,1,2,\dots \hat{=}$ S,P,D$,\dots$ of the underlying amplitudes $A_1$ and $A_2$,
\begin{equation}
    \label{eq:def-partial-wave-observable}
    S(\theta) \equiv F\left[A_1(\theta) A_2^*(\theta)\right] \equiv F\left[\left(\sum_{l_1=0}^\infty A_1^{(l_1)} p_{l_1}^{(m_1)}(\cos\theta)\right) \left(\sum_{l_2=0}^\infty A_2^{*(l_2)} p_{l_2}^{(m_2)}(\cos\theta)\right)\right]\,,
\end{equation}
where $F \in \{\text{Re},\text{Im}\}$ denotes taking either the real
or the imaginary part. Here $p_{l}^{(m)}$ denotes an \emph{associated Legendre} polynomial and $m_i$ is the third component of the angular momentum of the amplitude $A_i$, and we impose $m_1 \geq m_2$.\\

From the positivity of amplitudes, one immediately finds
\begin{equation}
    \int \rmdx{\cos\theta}  \, |A_i(\theta)|^2 = \sum_{l_i = 0}^\infty |A_i^{(l_i)}|^2 n_{l_i, m_i} < \infty\,,
\end{equation}
where we introduce $n_{l,m}$ via the scalar product of two associated Legendre polynomials,
\begin{equation}
    \label{eq:legendre-scalar-product}
    n_{l, m} \delta_{l, l'} \equiv \int_{-1}^1 \rmdx{\cos\theta} p_{l}^{(m)}(\cos\theta) p_{l'}^{(m)} (\cos\theta)  = \frac{2}{(2 l + 1)} \frac{(l + m)!}{(l - m)!} \delta_{l, l'}\,.
\end{equation}
Using the large-$l$ behaviour of $n_{l,m}$ one finds that with increasing $l_i$ the partial-wave amplitudes must asymptotically fall off as \fred{Let's say why. Unitarity?} \danny{Finiteness of observables, so yes: unitarity!}
\begin{equation}
    |A_i^{(l_i)}(\cos\theta)| \Big|_{l_i \gg m_i} \sim \frac{1}{\sqrt{l_i}}\,.
\end{equation}
This motivates to cut off the partial-wave expansion at some arbitrary angular momentum $L$.\\

Will will show in the following that such a cut off
is compatible with defining a basis of angular observables as coefficients of Legendre polynomials in $\cos\theta$. Since this expansion implies
a well defined total angular momentum for each observable, one ensures that the observables can in fact be disentangled experimentally.\\

@@

In the last step, we use Gaunt's formula \cite{Gaunt:1929} to integrate a triple product of associated Legendre polynomials,
\begin{equation}
\begin{aligned}
    T_{l_1,l_2,j}^{(m_1,m_2)}
        & = \int_{-1}^{+1} \rmdx{\cos\theta} p_{j}^{(m_1 + m_2)}(\cos\theta) p_{l_1}^{(m_1)}(\cos\theta) p_{l_2}^{(m_2)}(\cos\theta)\\
        & = (-1)^{s - l_1 - m_2} \frac{2 (l_1 + m_1)! (l_2 + m_2)! (2s - 2 l_2)! s!}{(l_1 - m_1)! (s - j)! (s - l_1)! (s - l_2)! (2s + 1)!}\\
        & \times \sum_{t=p}^q (-1)^t \frac{(j + m + t)!(l_1 + l_2 - m - t)!}{t! (j - m - t)! (l_1 - l_2 + m + t)! (l_2 - m_2 - t)!}\,,
\end{aligned}
\end{equation}
where
\begin{equation}
\begin{aligned}
    m & = m_1 + m_2\,, &
    m_1 & \geq m_2\,,  \\
    j, l_1, l_2 & \geq 0\,, &
    m, m_1, m_2 & \geq 0\,,
\end{aligned}
\end{equation}
and
\begin{equation}
\begin{aligned}
    s & = \frac{j + l_1 + l_2}{2}\,, &
    p & = \max(0, l_2 - l_1 - m)\,, &
    q & = \min(l_1 + l_2 - m, j - m, l_2 - m_2)\,.
\end{aligned}
\end{equation}
The necessary conditions for $T \neq 0$ are
\begin{equation}
    \label{eq:angular-momentum-addition}
    s \in \mathbb{N}\qquad \wedge \qquad l_1 - l_2 \leq l \leq l_1 + l_2,.
\end{equation}
The latter condition is well known from the addition rules of angular momenta. Note, however, that
the sum in \refeq{def-partial-wave-observable} goes to infinitely high angular momenta $l_1$ and $l_2$. As a consequence
of this and of \refeq{angular-momentum-addition}, the angular observables $P_{j,m}$
consist of sums with infinitely many terms. It is then up to theoretical analyses to
estimate or calculate the impact of individual contributions, with or without the cut off $L$.
For such an analysis in the decay $B\to K\pi\ell^+\ell^-$ we refer to \cite{Das:2014sra}.
We close by remarking that for any angular-momentum cut off $l < L$\fred{Do you mean $L < \infty$?}, one can find an upper limit for the squared sum of all remaining partial waves through a comparison with the inclusive decay
width.

>>>>

In several interesting processes we might only have an approximate
result for $P$.  In this section, we focus on one particular class of
mismodelling of the signal PDF: the angular-momentum cutoff in partial
wave expansions. This mismodelling potentially affects a large number
of decays and scattering processes. For the sake of clarity we take
the interesting class of four-body decays $B\to P_1 P_2 \ell_1
\bar\ell_2$ \fred{How general is this? The plot implies that the
  dimeson and the dilepton do not have a common vertex. But this
  requires an intermediate state, e.g., $K^{*}$}as an
example\footnote{This includes the rare $b\to s$ mediated $B$ decay $B
  \to K\pi\ell^+\ell^-$, and the $V_{ub}$ suppressed decay $B\to
  \pi\pi\ell^+\bar\nu_\ell$. For both examples the PDF $P$ is known in
  the small-width approximation and when assuming a pure $P$-wave,
  resonant final state. An extension to $S-P$ interference has been
  studied for $B\to K\pi\ell^+\ell^-$
  \cite{Blake:2012mb,Becirevic:2011bp}, and for $\bar{B}\to
  \pi\pi\ell^-\bar\nu_\ell$ \cite{Faller:2013dwa}. For a first study
  of $S$, $P$
  and $D$ interference, see \cite{Das:2014sra}}.\\

@@

\subsection{Recipe for \fred{Including} Detector Effects}
\label{sec:systematics:acceptance}

@@

In order to finally extract the angular observables from data, we use the \emph{measured raw moments}. Their MC estimator
$\lbrace \est{Q}_i\rbrace$ --- based on the \emph{detected events} $x^{(n)}$, $n=1,\dots,N$ --- reads
\begin{equation}
    \est{Q}_i \equiv R \frac{1}{N} \sum_n^N \tilde{f}_i(x^{(n)})\,.
\end{equation}
We then obtain MC estimators of the angular observables via
\begin{equation}
    \label{eq:estS}
    \est{S}_i \equiv \sum_{i} \left(\hat{M}^{-1}\right)_{ij} \est{Q}_j\,.
\end{equation}
The precise value of $R$ (see \refeq{def-R}) depends on the a priori
unknown angular observables $\lbrace S_i\rbrace$.  However, ratios and
other suitable combinations of the estimators $\lbrace
\est{S}_i\rbrace$ are independent of $R$. Conversely, $R$ can be
extracted from the self-consistency condition\footnote{This condition
  is process dependent; see \refeq{recipe:lambdabtolambdall} and
  \refeq{recipe:btokstarll}.}  $\hat{S}_0 = \frac{1}{2}$. Since $R$ is
relevant in the determination of the branching ratio, it is therefore
useful to carry out that determination and the extraction of the
angular observables in a combined analysis.

@@

\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/error-Q2_5_6_S5.png}
            \includegraphics[width=0.45\textwidth]{figs/error-Q2_5_6_S7.png}
            \caption{Uncertainty of the angular observables $S_5$ and $S_7$, extracted from simulated events of the decay $B\to K^*\ell^+\ell^-$ and as a function of the number of simulated events $K$. The red curve represents a fit to the function given in \refeq{unc-on-mean}.}
            \label{fig:errors}
            \fred{Plots convey exactly was is said in words. I see they are useful for a talk, but do we really need them in the write-up?}
\end{figure}

@@

Let us assume that the angular decomposition has been achieved for
some PDF $P$ for all angles except for one angle $\theta$. We now
focus on just one of the resulting observables and denote it by
$S(\theta)$. The dependence on the non-angular variables will be
ignored in the following. \fred{I rephrased it sligthly} Suppose $S$
has an expansion in terms of partial waves $l_1, l_2 = 0,1,2,\dots
\hat{=}$ S,P,D$,\dots$ of the underlying amplitudes $A_1$ and $A_2$,
\begin{equation}
    \label{eq:def-partial-wave-observable}
    S(\theta) \equiv F\left[A_1(\theta) A_2^*(\theta)\right] \equiv F\left[\left(\sum_{l_1=0}^\infty A_1^{(l_1)} p_{l_1}^{(m_1)}(\cos\theta)\right) \left(\sum_{l_2=0}^\infty A_2^{*(l_2)} p_{l_2}^{(m_2)}(\cos\theta)\right)\right]\,,
\end{equation}
where $F \in \{\text{Re},\text{Im}\}$ denotes taking either the real
or the imaginary part. Here $p_{l}^{(m)}$ denotes an \emph{associated Legendre} polynomial and $m_i$ is the third component of the angular momentum of the amplitude $A_i$, and we impose $m_1 \geq m_2$.\\

From the positivity of amplitudes, one immediately finds
\begin{equation}
    \int \rmdx{\cos\theta}  \, |A_i(\theta)|^2 = \sum_{l_i = 0}^\infty |A_i^{(l_i)}|^2 n_{l_i, m_i} < \infty\,,
\end{equation}
where we introduce $n_{l,m}$ via the scalar product of two associated Legendre polynomials,
\begin{equation}
    \label{eq:legendre-scalar-product}
    n_{l, m} \delta_{l, l'} \equiv \int_{-1}^1 \rmdx{\cos\theta} p_{l}^{(m)}(\cos\theta) p_{l'}^{(m)} (\cos\theta)  = \frac{2}{(2 l + 1)} \frac{(l + m)!}{(l - m)!} \delta_{l, l'}\,.
\end{equation}
Using the large-$l$ behaviour of $n_{l,m}$ one finds that with
increasing $l_i$ the partial-wave amplitudes must asymptotically fall
off as \fred{Let's say why. Unitarity?} \danny{Finiteness of
  observables, so yes: unitarity!} \fred{So can't we just say
  unitarity before the first equation? That's the reason why the sum
  need be finite, I guess. But if I look at (D2) again, it goes down as $1/(2l + 1)$, so If remember correctly, $A_i^{(l_i)} \propto 1/l^{\eps}$ would be enough to ensure convergence. So why $1/\sqrt{l_i}$?}
\begin{equation}
    |A_i^{(l_i)}(\cos\theta)| \Big|_{l_i \gg m_i} \sim \frac{1}{\sqrt{l_i}}\,.
\end{equation}
This motivates to cut off the partial-wave expansion at some arbitrary angular momentum $L$.\\

Will will show in the following that such a cutoff
is compatible with defining a basis of angular observables as coefficients of Legendre polynomials in $\cos\theta$. Since this expansion implies
a well defined total angular momentum for each observable, one ensures that the observables can in fact be disentangled experimentally.\\

@@

In the last step, we use Gaunt's formula \cite{Gaunt:1929} to integrate a triple product of associated Legendre polynomials,
\begin{equation}
\begin{aligned}
    T_{l_1,l_2,j}^{(m_1,m_2)}
        & = \int_{-1}^{+1} \rmdx{\cos\theta} p_{j}^{(m_1 + m_2)}(\cos\theta) p_{l_1}^{(m_1)}(\cos\theta) p_{l_2}^{(m_2)}(\cos\theta)\\
        & = (-1)^{s - l_1 - m_2} \frac{2 (l_1 + m_1)! (l_2 + m_2)! (2s - 2 l_2)! s!}{(l_1 - m_1)! (s - j)! (s - l_1)! (s - l_2)! (2s + 1)!}\\
        & \times \sum_{t=p}^q (-1)^t \frac{(j + m + t)!(l_1 + l_2 - m - t)!}{t! (j - m - t)! (l_1 - l_2 + m + t)! (l_2 - m_2 - t)!}\,,
\end{aligned}
\end{equation}
where
\begin{equation}
\begin{aligned}
    m & = m_1 + m_2\,, &
    m_1 & \geq m_2\,,  \\
    j, l_1, l_2 & \geq 0\,, &
    m, m_1, m_2 & \geq 0\,,
\end{aligned}
\end{equation}
and
\begin{equation}
\begin{aligned}
    s & = \frac{j + l_1 + l_2}{2}\,, &
    p & = \max(0, l_2 - l_1 - m)\,, &
    q & = \min(l_1 + l_2 - m, j - m, l_2 - m_2)\,.
\end{aligned}
\end{equation}
The necessary conditions for $T \neq 0$ are
\begin{equation}
    \label{eq:angular-momentum-addition}
    s \in \mathbb{N}\qquad \wedge \qquad l_1 - l_2 \leq l \leq l_1 + l_2,.
\end{equation}
The latter condition is well known from the addition rules of angular momenta. Note, however, that
the sum in \refeq{def-partial-wave-observable} goes to infinitely high angular momenta $l_1$ and $l_2$. As a consequence
of this and of \refeq{angular-momentum-addition}, the angular observables $P_{j,m}$
consist of sums with infinitely many terms. It is then up to theoretical analyses to
estimate or calculate the impact of individual contributions, with or without the cutoff $L$.
For such an analysis in the decay $B\to K\pi\ell^+\ell^-$ we refer to \cite{Das:2014sra}.
We close by remarking that for any angular-momentum cutoff $l < L$\fred{Do you mean $L < \infty$?}, one can find an upper limit for the squared sum of all remaining partial waves through a comparison with the inclusive decay
width.
